---
title: "R Notebook"
output: html_notebook
---

Render the notebook with commandline Rscript -e "rmarkdown::render('adjusting_covariates.Rmd')"

```{r loading packages}
library(readxl)
library(data.table)
library(stringr)
library(MASS)
library(parallel)
```

Reading beta and M values
```{r reading beta and M values}
# reading beta values
betaVal <- fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/qc_normlized_Beta_M/Beta_Values_qced_normlized.tsv", data.table = F)

#reading M values
mVal <-  fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/qc_normlized_Beta_M/M_Values_qced_normlized.tsv", data.table = F)
```

Reading Meta files
```{r reading meta information}
# read methylation meta info
meta_meth <- fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/raw_data/meta_meth.csv")

# read RNAseq meta info
meta_rnaseq <- fread("/sc/orga/projects/zhangb03a/shared/msbb-seq/Data/mixed_model_correction_updated/msbb.meta.BM_36.tsv")

#edit APOE and diagnosis format
meta_rnaseq$APOE <- gsub("/", "", gsub("e","", meta_rnaseq$APOE))
meta_rnaseq$APOE[is.na(meta_rnaseq$APOE)] <- "Unknown"
meta_rnaseq$Dx.by.braak.cerad[is.na(meta_rnaseq$Dx.by.braak.cerad)] <- "Unknown"
```

Mapping IDs between meta_meth and meta_rnaseq
```{r mapping IDs between meta_meth and meta_rnaseq}
#read id map file for methylation meta file
id_map_file <- read_xlsx("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/raw_data/MethylationBarcodeXSubNumXMinghuiXWang.xlsx", skip = 10)
setDT(id_map_file)
#pad the SubNum to 10 digits
id_map_file$SubNum <- str_pad(id_map_file$SubNum, 10, "left", 0)

id_map_file$SubNum <- paste0("AMPAD_MSSM_", id_map_file$SubNum)

#match IDs for meta_meth
meta_meth$SynapseID <- id_map_file$SubNum[match(meta_meth$Sample_Name, id_map_file$Sample_Name_Barcode)]

#ID mapping between meta_rnaseq & meta_meth
ID_overlap <- intersect(meta_meth$SynapseID, meta_rnaseq$SynapseId)

#subset meta_meth by ID_overlap
meta_meth <- meta_meth[SynapseID %in% ID_overlap]

# add sample ID column to meta_meth
meta_meth$SampleID <- paste(meta_meth$Sentrix_ID, meta_meth$Sentrix_Position, sep = "_")

# subset meta_rnaseq by ID_overlap
meta_rnaseq <- meta_rnaseq[SynapseId %in% ID_overlap]

# add sample ID column to meta_rnaseq
meta_rnaseq$SampleID <- meta_meth$SampleID[match(meta_rnaseq$SynapseId, meta_meth$SynapseID)]

# add another sample ID column to meta_rnaseq
meta_rnaseq$SampleID_2 <- id_map_file$Sample_ID[match(meta_rnaseq$SynapseId, id_map_file$SubNum)]

# add Sample batch column to meta_rnaseq
meta_rnaseq$Sentrix_ID <- meta_meth$Sentrix_ID[match(meta_rnaseq$SynapseId, meta_meth$SynapseID)]

meta_rnaseq$Sample_Source <- meta_meth$Sample_Source[match(meta_rnaseq$SynapseId, meta_meth$SynapseID)]
```


Removing samples with ambiguous sex and AD info
```{r}
meta_rnaseq <- meta_rnaseq[!SampleID_2 %in% c("PQDW544", "TVUF620", "KAST417", "DLUH936", "EOYS019") ]
```


Subsetting betaVal and mVal
```{r subsetting betaVal and mVal}
# subset betaVal
betaVal <- betaVal[, c("V1", meta_rnaseq$SampleID)]

# subset mVal
mVal <- mVal[, c("V1", meta_rnaseq$SampleID)]
```

Transposing betaVal and mVal
```{r transposing betaVal and mVal}
# for betaVal
rownames(betaVal) <- betaVal$V1
betaVal <- betaVal[, -1]
betaVal_t <- as.data.frame(t(betaVal))

# for mVal
rownames(mVal) <- mVal$V1
mVal <- mVal[, -1]
mVal_t <- as.data.frame(t(mVal))
```

Addding AOD, SEX_inferred, PMI, RACE_inferred, Sample_Source, columns to betaVal_t and mVal_t
```{r addding AOD, SEX_inferred, PMI, Sample_Source and RACE_inferred columns to betaVal_t and mVal_t}
# for betaVal
betaVal_t$PMI <- as.numeric(meta_rnaseq$PMI[match(rownames(betaVal_t), meta_rnaseq$SampleID)])
betaVal_t$RACE_inferred <- as.factor(meta_rnaseq$RACE_inferred[match(rownames(betaVal_t), meta_rnaseq$SampleID)])
betaVal_t$AOD <- as.numeric(meta_rnaseq$AOD[match(rownames(betaVal_t), meta_rnaseq$SampleID)])
betaVal_t$Sample_Source <- as.factor(meta_rnaseq$Sample_Source[match(rownames(betaVal_t), meta_rnaseq$SampleID)])
betaVal_t$SEX_inferred <- as.factor(meta_rnaseq$SEX_inferred[match(rownames(betaVal_t), meta_rnaseq$SampleID)])

# for mVal
mVal_t$PMI <- as.numeric(meta_rnaseq$PMI[match(rownames(mVal_t), meta_rnaseq$SampleID)])
mVal_t$RACE_inferred <- as.factor(meta_rnaseq$RACE_inferred[match(rownames(mVal_t), meta_rnaseq$SampleID)])
mVal_t$AOD <- as.numeric(meta_rnaseq$AOD[match(rownames(mVal_t), meta_rnaseq$SampleID)])
mVal_t$Sample_Source <- as.factor(meta_rnaseq$Sample_Source[match(rownames(mVal_t), meta_rnaseq$SampleID)])
mVal_t$SEX_inferred <- as.factor(meta_rnaseq$SEX_inferred[match(rownames(mVal_t), meta_rnaseq$SampleID)])
```


```{r calculating means for beta and m values}
# transpose betaVal
betaVal_t_numeric_cols <- 1:(ncol(betaVal_t)-5)
# calculate mean of all expression values
betaVal_t_mean <- mean(as.matrix(betaVal_t[, betaVal_t_numeric_cols]))


# transpose mVal
mVal_t_numeric_cols <- 1:(ncol(mVal_t)-5)
# calculate mean of all expression values
mVal_t_mean <- mean(as.matrix(mVal_t[, mVal_t_numeric_cols]))
```

Setting parallel clusters
```{r setting parallel clusters}
# Initiate cluster
cl <- parallel::makeCluster(16)

# load packages in each node
clusterEvalQ(cl, {
library(readxl)
library(data.table)
library(stringr)
library(MASS)
library(parallel)
})

# pass objects to each node
clusterExport(cl, c("ID_overlap", "betaVal", "betaVal_t", "betaVal_t_mean", "betaVal_t_numeric_cols", "mVal", "mVal_t", "mVal_t_mean", "mVal_t_numeric_cols", "id_map_file", "meta_meth", "meta_rnaseq"))
```

Adjusting for AOD, SEX_inferred, PMI, RACE_inferred, Sample_Source
```{r adjusting for AOD, SEX_inferred, PMI, RACE_inferred, Sample_Source}
# for betaVal
# regress out the covariates
betaVal_t_resid_age_sex <- parApply(cl, betaVal_t[, betaVal_t_numeric_cols], 2, function(x) rlm(x ~ betaVal_t[,"AOD"] + betaVal_t[,"SEX_inferred"] + betaVal_t[,"PMI"] + betaVal_t[,"RACE_inferred"] + betaVal_t[,"Sample_Source"])$resid)

# the adjusted expression values equal to the sum of the mean and residual
betaVal_t_resid_age_sex <- betaVal_t_resid_age_sex + betaVal_t_mean

# transpose the adjusted betaVal
betaVal_age_sex_adjusted <- as.data.frame(t(betaVal_t_resid_age_sex))



# for mVal
# regress out the covariates
mVal_t_resid_age_sex <- parApply(cl, mVal_t[, mVal_t_numeric_cols], 2, function(x) rlm(x ~ mVal_t[,"AOD"] + mVal_t[,"SEX_inferred"] + mVal_t[,"PMI"] + mVal_t[,"RACE_inferred"] + mVal_t[,"Sample_Source"])$resid)

# the adjusted expression values equal to the sum of the mean and residual
mVal_t_resid_age_sex <- mVal_t_resid_age_sex + mVal_t_mean

# transpose the adjusted mVal
mVal_age_sex_adjusted <- as.data.frame(t(mVal_t_resid_age_sex))
```


Adjusting for AOD, PMI, RACE_inferred, Sample_Source
```{r adjusting for AOD, PMI, RACE_inferred, Sample_Source}
# for betaVal
# regress out the covariates
betaVal_t_resid_age <- parApply(cl, betaVal_t[, betaVal_t_numeric_cols], 2, function(x) rlm(x ~ betaVal_t[,"AOD"] + betaVal_t[,"PMI"] + betaVal_t[,"RACE_inferred"] + betaVal_t[,"Sample_Source"])$resid)

# the adjusted expression values equal to the sum of the mean and residual
betaVal_t_resid_age <- betaVal_t_resid_age + betaVal_t_mean

# transpose the adjusted betaVal
betaVal_age_adjusted <- as.data.frame(t(betaVal_t_resid_age))



# for mVal
# regress out the covariates
mVal_t_resid_age <- parApply(cl, mVal_t[, mVal_t_numeric_cols], 2, function(x) rlm(x ~ mVal_t[,"AOD"] + mVal_t[,"PMI"] + mVal_t[,"RACE_inferred"] + mVal_t[,"Sample_Source"])$resid)

# the adjusted expression values equal to the sum of the mean and residual
mVal_t_resid_age <- mVal_t_resid_age + mVal_t_mean

# transpose the adjusted mVal
mVal_age_adjusted <- as.data.frame(t(mVal_t_resid_age))
```


Adjusting for SEX_inferred, PMI, RACE_inferred, Sample_Source
```{r adjusting for SEX_inferred, PMI, RACE_inferred, Sample_Source}
# for betaVal
# regress out the covariates
betaVal_t_resid_sex <- parApply(cl, betaVal_t[, betaVal_t_numeric_cols], 2, function(x) rlm(x ~ betaVal_t[,"SEX_inferred"] + betaVal_t[,"PMI"] + betaVal_t[,"RACE_inferred"] + betaVal_t[,"Sample_Source"])$resid)

# the adjusted expression values equal to the sum of the mean and residual
betaVal_t_resid_sex <- betaVal_t_resid_sex + betaVal_t_mean

# transpose the adjusted betaVal
betaVal_sex_adjusted <- as.data.frame(t(betaVal_t_resid_sex))



# for mVal
# regress out the covariates
mVal_t_resid_sex <- parApply(cl, mVal_t[, mVal_t_numeric_cols], 2, function(x) rlm(x ~ mVal_t[,"SEX_inferred"] + mVal_t[,"PMI"] + mVal_t[,"RACE_inferred"] + mVal_t[,"Sample_Source"])$resid)

# the adjusted expression values equal to the sum of the mean and residual
mVal_t_resid_sex <- mVal_t_resid_sex + mVal_t_mean

# transpose the adjusted mVal
mVal_sex_adjusted <- as.data.frame(t(mVal_t_resid_sex))
```


Adjusting for PMI, RACE_inferred, Sample_Source
```{r adjusting for PMI, RACE_inferred, Sample_Source}
# for betaVal
# regress out the covariates
betaVal_t_resid <- parApply(cl, betaVal_t[, betaVal_t_numeric_cols], 2, function(x) rlm(x ~ betaVal_t[,"PMI"] + betaVal_t[,"RACE_inferred"] + betaVal_t[,"Sample_Source"])$resid)

# the adjusted expression values equal to the sum of the mean and residual
betaVal_t_resid <- betaVal_t_resid + betaVal_t_mean

# transpose the adjusted betaVal
betaVal_adjusted <- as.data.frame(t(betaVal_t_resid))



# for mVal
# regress out the covariates
mVal_t_resid <- parApply(cl, mVal_t[, mVal_t_numeric_cols], 2, function(x) rlm(x ~ mVal_t[,"PMI"] + mVal_t[,"RACE_inferred"] + mVal_t[,"Sample_Source"])$resid)

# the adjusted expression values equal to the sum of the mean and residual
mVal_t_resid <- mVal_t_resid + mVal_t_mean

# transpose the adjusted mVal
mVal_adjusted <- as.data.frame(t(mVal_t_resid))
```


Wrting the adjusted mVals
```{r writing the adjusted mVals}
# writing the sex-age-adjusted mVals
fwrite(mVal_age_sex_adjusted, "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/qc_normlized_covariates_adjusted_Beta_M/MVals/M_Values_qced_normlized_plate_pmi_race_age_sex_adjusted.tsv", row.names = T, sep = "\t")

# writing the age-adjusted mVals
fwrite(mVal_age_adjusted, "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/qc_normlized_covariates_adjusted_Beta_M/MVals/M_Values_qced_normlized_plate_pmi_race_age_adjusted.tsv", row.names = T, sep = "\t")

# writing the sex-adjusted mVals
fwrite(mVal_sex_adjusted, "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/qc_normlized_covariates_adjusted_Beta_M/MVals/M_Values_qced_normlized_plate_pmi_race_sex_adjusted.tsv", row.names = T, sep = "\t")

# writing the adjusted mVals
fwrite(mVal_adjusted, "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/qc_normlized_covariates_adjusted_Beta_M/MVals/M_Values_qced_normlized_plate_pmi_race_adjusted.tsv", row.names = T, sep = "\t")
```

Wrting the adjusted betaVals
```{r writing the adjusted betaVals}
# writing the sex-age-adjusted betaVals
fwrite(betaVal_age_sex_adjusted, "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/qc_normlized_covariates_adjusted_Beta_M/BetaVals/Beta_Values_qced_normlized_plate_pmi_race_age_sex_adjusted.tsv", row.names = T, sep = "\t")

# writing the age-adjusted betaVals
fwrite(betaVal_age_adjusted, "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/qc_normlized_covariates_adjusted_Beta_M/BetaVals/Beta_Values_qced_normlized_plate_pmi_race_age_adjusted.tsv", row.names = T, sep = "\t")

# writing the sex-adjusted betaVals
fwrite(betaVal_sex_adjusted, "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/qc_normlized_covariates_adjusted_Beta_M/BetaVals/Beta_Values_qced_normlized_plate_pmi_race_sex_adjusted.tsv", row.names = T, sep = "\t")

# writing the adjusted betaVals
fwrite(betaVal_adjusted, "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/qc_normlized_covariates_adjusted_Beta_M/BetaVals/Beta_Values_qced_normlized_plate_pmi_race_adjusted.tsv", row.names = T, sep = "\t")
```
