---
title: "R Notebook"
output: html_notebook
---

Render the notebook with commandline Rscript -e "rmarkdown::render('make_expressionset.Rmd')"

```{r loading packages, message = F}
library(data.table)
library(readxl)
library(stringr)
library(limma)
library(Biobase)
```


Reading adjusted beta values
```{r reading batch pmi race adjusted beta values, message = F}
# reading batch pmi race adjusted beta values
invisible(betaVal_batch_pmi_race_adj <- fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/BetaVals/Beta_Values_qced_normlized_batch_pmi_race_adjusted.tsv", data.table = F))

# naming the rows with the probe column
rownames(betaVal_batch_pmi_race_adj) <- betaVal_batch_pmi_race_adj$V1
betaVal_batch_pmi_race_adj <- betaVal_batch_pmi_race_adj[, -1]

# convert betaVal_sex_adj to matrix
betaVal_batch_pmi_race_adj <- as.matrix(betaVal_batch_pmi_race_adj)
```

```{r reading batch pmi race age adjusted beta values, message = F}
# reading batch pmi race age adjusted beta values
invisible(betaVal_batch_pmi_race_age_adj <- fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/BetaVals/Beta_Values_qced_normlized_batch_pmi_race_age_adjusted.tsv", data.table = F))

# naming the rows with the probe column
rownames(betaVal_batch_pmi_race_age_adj) <- betaVal_batch_pmi_race_age_adj$V1
betaVal_batch_pmi_race_age_adj <- betaVal_batch_pmi_race_age_adj[, -1]

# convert betaVal_sex_adj to matrix
betaVal_batch_pmi_race_age_adj <- as.matrix(betaVal_batch_pmi_race_age_adj)
```

```{r reading batch pmi race sex adjusted beta values, message = F}
# reading batch pmi race sex adjusted beta values
invisible(betaVal_batch_pmi_race_sex_adj <- fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/BetaVals/Beta_Values_qced_normlized_batch_pmi_race_sex_adjusted.tsv", data.table = F))

# naming the rows with the probe column
rownames(betaVal_batch_pmi_race_sex_adj) <- betaVal_batch_pmi_race_sex_adj$V1
betaVal_batch_pmi_race_sex_adj <- betaVal_batch_pmi_race_sex_adj[, -1]

# convert betaVal_sex_adj to matrix
betaVal_batch_pmi_race_sex_adj <- as.matrix(betaVal_batch_pmi_race_sex_adj)
```

```{r reading batch pmi race sex age adjusted beta values, message = F}
# reading batch pmi race sex age adjusted beta values
invisible(betaVal_batch_pmi_race_age_sex_adj <- fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/BetaVals/Beta_Values_qced_normlized_batch_pmi_race_age_sex_adjusted.tsv", data.table = F))

# naming the rows with the probe column
rownames(betaVal_batch_pmi_race_age_sex_adj) <- betaVal_batch_pmi_race_age_sex_adj$V1
betaVal_batch_pmi_race_age_sex_adj <- betaVal_batch_pmi_race_age_sex_adj[, -1]

# convert betaVal_sex_adj to matrix
betaVal_batch_pmi_race_age_sex_adj <- as.matrix(betaVal_batch_pmi_race_age_sex_adj)
```


Reading adjusted M values
```{r reading batch pmi race adjusted M values, message = F}
# reading batch pmi race adjusted M values
invisible(MVal_batch_pmi_race_adj <- fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/MVals/M_Values_qced_normlized_batch_pmi_race_adjusted.tsv", data.table = F))

# naming the rows with the probe column
rownames(MVal_batch_pmi_race_adj) <- MVal_batch_pmi_race_adj$V1
MVal_batch_pmi_race_adj <- MVal_batch_pmi_race_adj[, -1]

# convert MVal_sex_adj to matrix
MVal_batch_pmi_race_adj <- as.matrix(MVal_batch_pmi_race_adj)
```

```{r reading batch pmi race age adjusted M values, message = F}
# reading batch pmi race age adjusted M values
invisible(MVal_batch_pmi_race_age_adj <- fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/MVals/M_Values_qced_normlized_batch_pmi_race_age_adjusted.tsv", data.table = F))

# naming the rows with the probe column
rownames(MVal_batch_pmi_race_age_adj) <- MVal_batch_pmi_race_age_adj$V1
MVal_batch_pmi_race_age_adj <- MVal_batch_pmi_race_age_adj[, -1]

# convert MVal_sex_adj to matrix
MVal_batch_pmi_race_age_adj <- as.matrix(MVal_batch_pmi_race_age_adj)
```

```{r reading batch pmi race sex adjusted M values, message = F}
# reading batch pmi race sex adjusted M values
invisible(MVal_batch_pmi_race_sex_adj <- fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/MVals/M_Values_qced_normlized_batch_pmi_race_sex_adjusted.tsv", data.table = F))

# naming the rows with the probe column
rownames(MVal_batch_pmi_race_sex_adj) <- MVal_batch_pmi_race_sex_adj$V1
MVal_batch_pmi_race_sex_adj <- MVal_batch_pmi_race_sex_adj[, -1]

# convert MVal_sex_adj to matrix
MVal_batch_pmi_race_sex_adj <- as.matrix(MVal_batch_pmi_race_sex_adj)
```

```{r reading batch pmi race sex age adjusted M values, message = F}
# reading batch pmi race sex age adjusted M values
invisible(MVal_batch_pmi_race_age_sex_adj <- fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/MVals/M_Values_qced_normlized_batch_pmi_race_age_sex_adjusted.tsv", data.table = F))

# naming the rows with the probe column
rownames(MVal_batch_pmi_race_age_sex_adj) <- MVal_batch_pmi_race_age_sex_adj$V1
MVal_batch_pmi_race_age_sex_adj <- MVal_batch_pmi_race_age_sex_adj[, -1]

# convert MVal_sex_adj to matrix
MVal_batch_pmi_race_age_sex_adj <- as.matrix(MVal_batch_pmi_race_age_sex_adj)
```


Reading Meta files
```{r reading meta information}
# read methylation meta info
meta_meth <- fread("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/raw_data/meta_meth.csv")

# read RNAseq meta info
meta_rnaseq <- fread("/sc/orga/projects/zhangb03a/shared/msbb-seq/Data/mixed_model_correction_updated/msbb.meta.BM_36.tsv")

#edit APOE and diagnosis format
meta_rnaseq$APOE <- gsub("/", "", gsub("e","", meta_rnaseq$APOE))
meta_rnaseq$APOE[is.na(meta_rnaseq$APOE)] <- "Unknown"
meta_rnaseq$Dx.by.braak.cerad[is.na(meta_rnaseq$Dx.by.braak.cerad)] <- "Unknown"

#add new group columns
meta_rnaseq[, c("sex.ad.apoe", "ad.apoe", "sex.ad") := list(paste(SEX_inferred, Dx.by.braak.cerad, APOE, sep = "."), paste(Dx.by.braak.cerad, APOE, sep = "."), paste(SEX_inferred, Dx.by.braak.cerad, sep = "."))]
```

Mapping IDs between meta_meth and meta_rnaseq
```{r mapping IDs between meta_meth and meta_rnaseq}
#read id map file for methylation meta file
id_map_file <- read_xlsx("/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/raw_data/MethylationBarcodeXSubNumXMinghuiXWang.xlsx", skip = 10)
setDT(id_map_file)
#pad the SubNum to 10 digits
id_map_file$SubNum <- str_pad(id_map_file$SubNum, 10, "left", 0)

id_map_file$SubNum <- paste0("AMPAD_MSSM_", id_map_file$SubNum)

#match IDs for meta_meth
meta_meth$SynapseID <- id_map_file$SubNum[match(meta_meth$Sample_Name, id_map_file$Sample_Name_Barcode)]

#ID mapping between meta_rnaseq & meta_meth
ID_overlap <- intersect(meta_meth$SynapseID, meta_rnaseq$SynapseId)

#subset meta_meth by ID_overlap
meta_meth <- meta_meth[SynapseID %in% ID_overlap]

# add sample ID column to meta_meth
meta_meth$SampleID <- paste(meta_meth$Sentrix_ID, meta_meth$Sentrix_Position, sep = "_")

# subset meta_rnaseq by ID_overlap
meta_rnaseq <- meta_rnaseq[SynapseId %in% ID_overlap]

# add sample ID column to meta_rnaseq
meta_rnaseq$SampleID <- meta_meth$SampleID[match(meta_rnaseq$SynapseId, meta_meth$SynapseID)]

# add another sample ID column to meta_rnaseq
meta_rnaseq$SampleID_2 <- id_map_file$Sample_ID[match(meta_rnaseq$SynapseId, id_map_file$SubNum)]

# add Sample batch column to meta_rnaseq
meta_rnaseq$Sentrix_ID <- meta_meth$Sentrix_ID[match(meta_rnaseq$SynapseId, meta_meth$SynapseID)]
```


Removing samples with ambiguous sex and AD info
```{r}
meta_rnaseq <- meta_rnaseq[!SampleID_2 %in% c("PQDW544", "TVUF620", "KAST417", "DLUH936", "EOYS019") ]
```


Matching sample order between meta_rnaseq and beta/m matricies
```{r matching sample order between meta_rnaseq and beta/m matricies}
# check if column order is same for all beta matricies
sapply(list(colnames(betaVal_batch_pmi_race_adj), colnames(betaVal_batch_pmi_race_age_adj), colnames(betaVal_batch_pmi_race_sex_adj), colnames(MVal_batch_pmi_race_adj), colnames(MVal_batch_pmi_race_age_adj), colnames(MVal_batch_pmi_race_sex_adj), colnames(MVal_batch_pmi_race_age_sex_adj)), identical, colnames(betaVal_batch_pmi_race_age_sex_adj))

# match meta_rnaseq and betaVal_batch_pmi_race_adj sample order
meta_rnaseq <- meta_rnaseq[match(colnames(betaVal_batch_pmi_race_adj), meta_rnaseq$SampleID), ]
setDF(meta_rnaseq)
rownames(meta_rnaseq) <- meta_rnaseq$SampleID
```


Making expressionset for beta values
```{r assembling expressionset for betaVal_batch_pmi_race_adj}
# check if rownames and colnames of meta_rnaseq and betaVal_batch_pmi_race_adj are same
identical(colnames(betaVal_batch_pmi_race_adj), rownames(meta_rnaseq))

# create meta info for the expressionset
phenoData_betaVal_batch_pmi_race_adj <- new("AnnotatedDataFrame", data = meta_rnaseq)

# construct an expressionset for betaVal_batch_pmi_race_adj
exprset_betaVal_batch_pmi_race_adj <- ExpressionSet(assayData = betaVal_batch_pmi_race_adj, phenoData = phenoData_betaVal_batch_pmi_race_adj)
```

```{r assembling expressionset for betaVal_batch_pmi_race_age_adj}
# check if rownames and colnames of meta_rnaseq and betaVal_batch_pmi_race_age_adj are same
identical(colnames(betaVal_batch_pmi_race_age_adj), rownames(meta_rnaseq))

# create meta info for the expressionset
phenoData_betaVal_batch_pmi_race_age_adj <- new("AnnotatedDataFrame", data = meta_rnaseq)

# construct an expressionset for betaVal_batch_pmi_race_age_adj
exprset_betaVal_batch_pmi_race_age_adj <- ExpressionSet(assayData = betaVal_batch_pmi_race_age_adj, phenoData = phenoData_betaVal_batch_pmi_race_age_adj)
```

```{r assembling expressionset for betaVal_batch_pmi_race_age_sex_adj}
# check if rownames and colnames of meta_rnaseq and betaVal_batch_pmi_race_age_sex_adj are same
identical(colnames(betaVal_batch_pmi_race_age_sex_adj), rownames(meta_rnaseq))

# create meta info for the expressionset
phenoData_betaVal_batch_pmi_race_age_sex_adj <- new("AnnotatedDataFrame", data = meta_rnaseq)

# construct an expressionset for betaVal_batch_pmi_race_age_sex_adj
exprset_betaVal_batch_pmi_race_age_sex_adj <- ExpressionSet(assayData = betaVal_batch_pmi_race_age_sex_adj, phenoData = phenoData_betaVal_batch_pmi_race_age_sex_adj)
```

```{r assembling expressionset for betaVal_batch_pmi_race_sex_adj}
# check if rownames and colnames of meta_rnaseq and betaVal_batch_pmi_race_sex_adj are same
identical(colnames(betaVal_batch_pmi_race_sex_adj), rownames(meta_rnaseq))

# create meta info for the expressionset
phenoData_betaVal_batch_pmi_race_sex_adj <- new("AnnotatedDataFrame", data = meta_rnaseq)

# construct an expressionset for betaVal_batch_pmi_race_sex_adj
exprset_betaVal_batch_pmi_race_sex_adj <- ExpressionSet(assayData = betaVal_batch_pmi_race_sex_adj, phenoData = phenoData_betaVal_batch_pmi_race_sex_adj)
```



Making expressionset for M values
```{r assembling expressionset for MVal_batch_pmi_race_adj}
# check if rownames and colnames of meta_rnaseq and MVal_batch_pmi_race_adj are same
identical(colnames(MVal_batch_pmi_race_adj), rownames(meta_rnaseq))

# create meta info for the expressionset
phenoData_MVal_batch_pmi_race_adj <- new("AnnotatedDataFrame", data = meta_rnaseq)

# construct an expressionset for MVal_batch_pmi_race_adj
exprset_MVal_batch_pmi_race_adj <- ExpressionSet(assayData = MVal_batch_pmi_race_adj, phenoData = phenoData_MVal_batch_pmi_race_adj)
```

```{r assembling expressionset for MVal_batch_pmi_race_age_adj}
# check if rownames and colnames of meta_rnaseq and MVal_batch_pmi_race_age_adj are same
identical(colnames(MVal_batch_pmi_race_age_adj), rownames(meta_rnaseq))

# create meta info for the expressionset
phenoData_MVal_batch_pmi_race_age_adj <- new("AnnotatedDataFrame", data = meta_rnaseq)

# construct an expressionset for MVal_batch_pmi_race_age_adj
exprset_MVal_batch_pmi_race_age_adj <- ExpressionSet(assayData = MVal_batch_pmi_race_age_adj, phenoData = phenoData_MVal_batch_pmi_race_age_adj)
```

```{r assembling expressionset for MVal_batch_pmi_race_age_sex_adj}
# check if rownames and colnames of meta_rnaseq and MVal_batch_pmi_race_age_sex_adj are same
identical(colnames(MVal_batch_pmi_race_age_sex_adj), rownames(meta_rnaseq))

# create meta info for the expressionset
phenoData_MVal_batch_pmi_race_age_sex_adj <- new("AnnotatedDataFrame", data = meta_rnaseq)

# construct an expressionset for MVal_batch_pmi_race_age_sex_adj
exprset_MVal_batch_pmi_race_age_sex_adj <- ExpressionSet(assayData = MVal_batch_pmi_race_age_sex_adj, phenoData = phenoData_MVal_batch_pmi_race_age_sex_adj)
```

```{r assembling expressionset for MVal_batch_pmi_race_sex_adj}
# check if rownames and colnames of meta_rnaseq and MVal_batch_pmi_race_sex_adj are same
identical(colnames(MVal_batch_pmi_race_sex_adj), rownames(meta_rnaseq))

# create meta info for the expressionset
phenoData_MVal_batch_pmi_race_sex_adj <- new("AnnotatedDataFrame", data = meta_rnaseq)

# construct an expressionset for MVal_batch_pmi_race_sex_adj
exprset_MVal_batch_pmi_race_sex_adj <- ExpressionSet(assayData = MVal_batch_pmi_race_sex_adj, phenoData = phenoData_MVal_batch_pmi_race_sex_adj)
```


Writing expressionsets
```{r writing expressionsets for beta values}
saveRDS(exprset_betaVal_batch_pmi_race_adj, file = "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/BetaVals_expressionset/exprset_Beta_Values_qced_normlized_batch_pmi_race_adjusted.rds")

saveRDS(exprset_betaVal_batch_pmi_race_age_adj, file = "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/BetaVals_expressionset/exprset_Beta_Values_qced_normlized_batch_pmi_race_age_adjusted.rds")

saveRDS(exprset_betaVal_batch_pmi_race_age_sex_adj, file = "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/BetaVals_expressionset/exprset_Beta_Values_qced_normlized_batch_pmi_race_age_sex_adjusted.rds")

saveRDS(exprset_betaVal_batch_pmi_race_sex_adj, file = "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/BetaVals_expressionset/exprset_Beta_Values_qced_normlized_batch_pmi_race_sex_adjusted.rds")
```

```{r writing expressionsets for M values}
saveRDS(exprset_MVal_batch_pmi_race_adj, file = "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/MVals_expressionset/exprset_M_Values_qced_normlized_batch_pmi_race_adjusted.rds")

saveRDS(exprset_MVal_batch_pmi_race_age_adj, file = "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/MVals_expressionset/exprset_M_Values_qced_normlized_batch_pmi_race_age_adjusted.rds")

saveRDS(exprset_MVal_batch_pmi_race_age_sex_adj, file = "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/MVals_expressionset/exprset_M_Values_qced_normlized_batch_pmi_race_age_sex_adjusted.rds")

saveRDS(exprset_MVal_batch_pmi_race_sex_adj, file = "/sc/orga/projects/zhangb03a/lei_guo/BM36_methylation_analysis/processed_data/covariates_adjusted_Beta_M/MVals_expressionset/exprset_M_Values_qced_normlized_batch_pmi_race_sex_adjusted.rds")
```

